New lower bounds for Schur and weak Schur numbers

### 2021-04-09 article/weakschur.tex
premiere version (pas de relecture + mise en page)

@@

Note that with this numbering of columns, the column of the sum of two numbers is the only integer in \([\![-l,q]\!]\) equal to two the sum of the columns modulo \(a\).

\renewcommand{\arraystretch}{2}
\begin{center}
\[
\begin{array}{|c|c|c|c|c|c|c|c|c|>{\columncolor{blue}}c|>{\columncolor{blue}}c|>{\columncolor{blue}}c|>{\columncolor{green}} c|}
\cline{6-9}
   \multicolumn{5}{c|}{} & \cellcolor{green}1 & \cellcolor{green}2 & \cellcolor{blue} 3&\cellcolor{green}4&5&6&7&8 \\
\hline
   \cellcolor{red}9 & \cellcolor{red}10 & \cellcolor{red}11 & \cellcolor{red}12 &\cellcolor{red}13&\cellcolor{red}14&\cellcolor{red}15&\cellcolor{red}16&\cellcolor{red}17&18&19&20&21\\
\hline
\cellcolor{yellow}22&\cellcolor{yellow}23&\cellcolor{yellow}24&\cellcolor{yellow}25&\cellcolor{yellow}26&\cellcolor{yellow}27&\cellcolor{yellow}28&\cellcolor{yellow}29&\cellcolor{yellow}30&31&32&33&34\\
\hline
...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...&...&...&...\\
\hline
\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&...&...&...&...\\
\hline
...&...&...&...&...&...&...&...&...&...&...&...&...\\
\hline
\end{array}
\]
\end{center}

@@

We denote by \(f\) the projection of the equivalence relation induced by the partition of \([\![1,q]\!]\) and \(g\) the
one induced by the partition of \([\![1,p]\!]\). Each equivalence class is represented by a single integer, therefore :

@@

@@

\begin{remark}
In the following theorem, \(\lambda\) is the function which maps an element x to its line number.
\end{remark}

@@

\begin{definition}
Let \( (p,k,b) \in (\mathbb{N}^*)^3\), Let \((A_1,....,A_k)\) a partition of  \([\![1, p]\!]\).
This partition is said to be a b-weakly-sum-free template (b-WSF-template) of \(k\) colors and lenght \(p\) when:
\\\\
\underline{\(\forall i \in [\![1, k]\!], \quad A_i\) is weakly-sum-free}
\\\\
\underline{\(\forall i \in [\![1, k]\!], \quad A_i\backslash [\![1, b]\!]\) is sum-free}
\\\\
\underline{For \(A_k\) (the special subset):} \quad \(\forall (x,y) \in A_k^2,\)
\\
\[if \quad x+y>b+2(p-b),\quad x+y-2(p-b)\notin A_k\]
\\
\underline{For the others subsets:}\quad \(\forall i \in [\![0,k-1]\!], \forall(x,y) \in A_i^2\)
\\
\[
x+y>p \implies \pi(x+y) \notin A_i
\]
\end{definition}

@@

\textsc{Proof :}Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\), let a=q-b. \\
We denote by \(g\) the projection of the equivalence relation induced by the partition of \([\![1,q]\!]\) and \(h\) the
one induced by the partition of \([\![1,p]\!]\). Each equivalence class is represented by a single integer, therefore :
\[ g : [\![1,q]\!] \longrightarrow [\![1,n]\!] \text{ and } (A_{g^{-1}(1)},...A_{g^{-1}(q)})\text{ is a b-WSF-template.}
\]
\[h : [\![1,p]\!] \longrightarrow [\![1,k]\!] \text{ and } \forall (x,y) \in [\![1,q]\!]^2 \text{, } h(x) = h(y)
\Longrightarrow h(x+y) \neq h(x)
\]

Let \( f : [\![1,b+pa]\!] \longrightarrow [\![1,n]\!] \) such as:
\\\\-if x \(\leqslant b \text{ (we will note x} \in \mathcal{T}): f(x)=g(x)\)
\\\\-if \( x \in [\![1,b+pa]\!] \text{ and } \pi(x) \notin A_n \text{ (we will note x} \in \mathcal{C}): f(x)=g(\pi(x))\)
\\\\-if \( x \in [\![1,b+pa]\!]  \text{ and }\pi(x) \in A_n \text{ (we will note x} \in \mathcal{L}): f(x)=n-1+h(\lambda(x))\)
\\\\
f is well defined because \(\pi\) is defined for x\(>b\) and \(\forall x \in \![1,b+pa]\!], f(x)\leqslant n+k-1\) because h(\(\lambda(x))\leqslant k\)

@@

\subsection{New lower bounds for Weak Schur numbers}
Having found suitable templates, which can be found in the appendix, with a SAT solver, we claim that for all \(n \in
\mathbb{N}^*\):
\[
4S(n) + 1 \leqslant WS(n+1)
\]
\[
13S(n) + 8 \leqslant WS(n+2)
\]
\[
42S(n) + 24 \leqslant WS(n+3)
\]
\[
127S(n) + 68 \leqslant WS(n+4)
\]
The first two inequalities are due to Rowley, they are detailed in [2]. Like in 3.3, we compute the lower bounds given
by the previous inequalities for \( n \in [\![8,15]\!] \). The best lower bound for each integer is highlighted.\\
\\
\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	n & 8 & 9 & 10 & 11 \\
	\hline
	\(4S(n-1) + 2 \) & 6722 & 21146 & \cellcolor{yellow} 71214 & \cellcolor{yellow} 243794\\
	\hline
	\(13S(n-2) + 8 \) & \cellcolor{yellow} 6976 & 21848 & 68726 & 231447\\
	\hline
	\(42S(n-3) + 24 \) & 6744 & \cellcolor{yellow} 22536 & 70584 & 222036 \\
	\hline
	\(127S(n-4) + 68 \) & 5656 & 20388 & 68140 & 213428\\
	\hline
	\hline
	n & 12 & 13 & 14 & 15 \\
	\hline
	\(4S(n-1) + 2 \) & \cellcolor{yellow} 815314 & 2554194 & 8045162 & \cellcolor{yellow} 27061154\\
	\hline
	\(13S(n-2) + 8 \) & 792332 & \cellcolor{yellow}2649772 & 8301132 & 26146778 \\
	\hline
	\(42S(n-3) + 24 \) & 747750 & 2559840 & \cellcolor{yellow} 8560800 &  25886224 \\
	\hline
	\(127S(n-4) + 68 \) & 671390 & 2261049 & 7740464 & 25886224 \\
	\hline
\end{tabular}
\end{center}

With \( S(9) \geqslant 17803 \), we found a new lower bound for
\(WS(10)\) using Rowley's inequality. Moreover, the third inequality
gives new lower bounds for \(WS(9)\) and \(WS(14)\). However, the last
inequality doesn't give any better lower bound, even beyond \( n = 15 \)
: the best bounds are always provided by the first three. We highly suspect that these values can be improved by
investigating the search
space further, which would provide new, more effective templates. One
may try to go over this search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP},
but with a different search space (as we explain in \hyperref[SAT]{SAT section}).
This could be the suject of a future work.

>>>>

\begin{remark}
This can be seen as an equivalent for weak Schur numbers of Abott and Hanson's construction for Schur numbers.
\end{remark}

@@

Like for SF-templates, \(\mathcal{L}\) and \(\mathcal{C}\) play the role of security zones for each other. Note that with 
this numbering of columns, the column of the sum of two numbers is the only integer in \([\![-l,q]\!]\) equal to two the 
sum of the columns modulo \(a\).

\renewcommand{\arraystretch}{2}
\begin{center}
\[
\begin{array}{|c|c|c|c|c|c|c|c|c|>{\columncolor{blue}}c|>{\columncolor{blue}}c|>{\columncolor{blue}}c|>{\columncolor{green}} c|}
\cline{6-9}
   \multicolumn{5}{c|}{} & \cellcolor{green} 1 & \cellcolor{green} 2 & \cellcolor{blue} ... &\cellcolor{green} r & r + 1 & ... & b - 1 &  b \\
\hline
   \cellcolor{red} a - l & \cellcolor{red} a- l + 1 & \cellcolor{red} ... & \cellcolor{red} a - 1 & \cellcolor{red} a & \cellcolor{red} a + 1 & \cellcolor{red} ... &\cellcolor{red} a + r - 1 & \cellcolor{red} a + r & a + r + 1 & ... & a + b - 1 & a+ b \\
\hline
\cellcolor{yellow} 2 a - l &\cellcolor{yellow}...&\cellcolor{yellow}...&\cellcolor{yellow}...&\cellcolor{yellow} 2 a &\cellcolor{yellow} ... &\cellcolor{yellow} ... &\cellcolor{yellow} ... &\cellcolor{yellow} 2 a + r & ... & ... & ... & 2 a + b \\
\hline
...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...\cellcolor{yellow}&...&...&...&...\\
\hline
\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&\cellcolor{red}...&...&...&...&...\\
\hline
p a - l & ... & ... & ... &p a & ... & ... & .... & p a + r & ... & ... & ... & p a + b\\
\hline
\end{array}
\]
\end{center}

@@

We denote by \(f\) the colouring associated to the partition of \([\![1,q]\!]\) and \(g\) the
one associated to the partition of \([\![1,p]\!]\).

@@

\begin{remark}
\(\pi\) is the projection on the first line mentioned in the intuitive explanation.
\end{remark}

@@

\begin{remark}
In the following theorem, \(\lambda\) is the function which maps an element x to its line number, as mentioned in the intuitive explanation.
\end{remark}

@@

\begin{definition}
Let \( (p,n,b) \in (\mathbb{N}^*)^3\), Let \((A_1,...,A_n)\) a partition of  \([\![1, p]\!]\).
This partition is said to be a b-weakly-sum-free template (b-WSF-template) of \(n\) colors and lenght \(p\) when:
\\\\
\underline{\(\forall i \in [\![1, n]\!], \quad A_i\) is weakly-sum-free}
\\\\
\underline{\(\forall i \in [\![1, n]\!], \quad A_i\backslash [\![1, b]\!]\) is sum-free}
\\\\
\underline{For \(A_n\) (the special subset):} \quad \(\forall (x,y) \in A_n^2,\)
\\
\[if \quad x+y>b+2(p-b),\quad x+y-2(p-b)\notin A_n\]
\\
\underline{For the others subsets:}\quad \(\forall i \in [\![0,n-1]\!], \forall(x,y) \in A_i^2,\)
\\
\[
x+y>p \implies \pi(x+y) \notin A_i
\]
\end{definition}

\begin{remark}
	Please note that the special color \(n\) is not necessarily the last color by order of appearance.
\end{remark}

@@

\textsc{Proof :}Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\), let a=q-b. \\
We denote by \(g\) the colouring associated to the partition of \([\![1,q]\!]\) and \(h\) the
one associated to the partition of \([\![1,p]\!]\).
\[ g : [\![1,q]\!] \longrightarrow [\![1,n]\!] \text{ and } (A_{g^{-1}(1)},...A_{g^{-1}(q)})\text{ is a b-WSF-template.}
\]
\[h : [\![1,p]\!] \longrightarrow [\![1,k]\!] \text{ and } \forall (x,y) \in [\![1,q]\!]^2 \text{, } h(x) = h(y)
\Longrightarrow h(x+y) \neq h(x)
\]

Define \( f : [\![1,b+pa]\!] \longrightarrow [\![1,n]\!] \) such that:
\\\\-if x \(\leqslant b \text{ (we will note x} \in \mathcal{T}): f(x)=g(x)\)
\\\\-if \( x \in [\![1,b+pa]\!] \text{ and } \pi(x) \notin A_n \text{ (we will note x} \in \mathcal{C}): f(x)=g(\pi(x))\)
\\\\-if \( x \in [\![1,b+pa]\!]  \text{ and }\pi(x) \in A_n \text{ (we will note x} \in \mathcal{L}): f(x)=n-1+h(\lambda(x))\)
\\\\
f is well defined because \(\pi\) is defined for x\(>b\) and \(\forall x \in \![1,b+pa]\!], f(x)\leqslant n+k-1\) because h(\(\lambda(x))\leqslant k\)

@@

\begin{corollary}
The general lower bound for weak Schur numbers in function of both regular and weak Schur numbers can be seen as a particular 
case of WSF-template in the same way Abott and Hanson's construction can be seen as a particular case of SF-template.
\end{corollary}

\subsection{New lower bounds for Weak Schur numbers}
Having found suitable templates, which can be found in the appendix, with a SAT solver, we claim that for all \(n \in
\mathbb{N}^*\):

\[
4S(n) + 1 \leqslant WS(n+1)
\]
\[
13S(n) + 8 \leqslant WS(n+2)
\]
\[
42S(n) + 24 \leqslant WS(n+3)
\]
\[
127S(n) + 68 \leqslant WS(n+4)
\]

The first two inequalities were found by Rowley, they are detailed in [2]. The third inequality is optimal and was found with a SAT solver. 
As for the fourth inequality, it is not useful since by comibing the first inequality with \(S(n) \geqslant 33 S(n-3) + 6\) provides a 
better inequality. It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of 
5 colors which start with a good \(WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates 
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates 
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}. 
This could be the suject of a future work.

Like in 3.3, we compute the lower bounds given by the previous inequalities for \( n \in [\![8,15]\!] \). The best lower bound 
for each integer is highlighted.\\
\\
\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	n & 8 & 9 & 10 & 11 \\
	\hline
	\(4S(n-1) + 2 \) & 6722 & 21146 & \cellcolor{yellow} 71214 & \cellcolor{yellow} 243794\\
	\hline
	\(13S(n-2) + 8 \) & \cellcolor{yellow} 6976 & 21848 & 68726 & 231447\\
	\hline
	\(42S(n-3) + 24 \) & 6744 & \cellcolor{yellow} 22536 & 70584 & 222036 \\
	\hline
	\(127S(n-4) + 68 \) & 5656 & 20388 & 68140 & 213428\\
	\hline
	\hline
	n & 12 & 13 & 14 & 15 \\
	\hline
	\(4S(n-1) + 2 \) & \cellcolor{yellow} 815314 & 2554194 & 8045162 & \cellcolor{yellow} 27061154\\
	\hline
	\(13S(n-2) + 8 \) & 792332 & \cellcolor{yellow}2649772 & 8301132 & 26146778 \\
	\hline
	\(42S(n-3) + 24 \) & 747750 & 2559840 & \cellcolor{yellow} 8560800 &  25886224 \\
	\hline
	\(127S(n-4) + 68 \) & 671390 & 2261049 & 7740464 & 25886224 \\
	\hline
\end{tabular}
\end{center}

With \( S(9) \geqslant 17803 \), we found a new lower bound for \(WS(10)\) using Rowley's inequality. 
Moreover, the third inequality gives new lower bounds for \(WS(9)\) and \(WS(14)\).
